Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography

### 2024-05-22 manuscript/manuscript.tex
Final changes before publication

\authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2024). Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography. \textit{PsyArXiv}. Version~3. \url{https://doi.org/10.31234/osf.io/qnz5k}}

@@

Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation. All data and code is available from \url{https://osf.io/7auw6/}. \firstrevision{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}. Note, however, that due to the more exploratory nature of this project, our preregistration did not specify strong confirmatory hypotheses or precise statistical models, focusing instead on the research question, experimental conditions, general predictions, primary measurement constructs, sample size, and exclusion criteria.}

@@

Like Experiment~1, redundant systems are characteristic of the Transmission-only condition, especially in Epoch~III. As the spoken suffixes become more homophonous, the variant spellings are increasingly conditioned on shape rather than color, perhaps because the spoken language signals to learners that the language does not mark color, so they rationalize the system as three words with idiosyncratic spellings. Interestingly, however, all chains exhibited conservation of spelling \textit{form}, even if the way in which spelling was conditioned on meaning was lost. For example, Chain~A ultimately represents the sound /fə/ with the spellings <foe> and <xa>, spellings that are internally inconsistent and contrary to standard uses of the Roman alphabet, but which trace their origins back to the original seed orthography. Overall, then, the Transmission-only condition in Experiment~2 is characterized by the conservation of spelling \textit{form} without conserving how form patterns with meaning.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig12.pdf}}
	\vspace*{2pt}
	\caption{Communicative success by generation in the communicative conditions of all three experiments. The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3).}
	\label{fig12}
	\end{figure}

>>>>

\authornote{Carr, J.\@ W., \& Rastle, K.\@ (2024). Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography. \textit{Cognition}, \textit{249}, Article~105809. \url{https://doi.org/10.1016/j.cognition.2024.105809}}

@@

Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation. All data and code is available from \url{https://osf.io/7auw6/}. \firstrevision{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/p8aw9.pdf}. Note, however, that due to the more exploratory nature of this project, our preregistration did not specify strong confirmatory hypotheses or precise statistical models, focusing instead on the research question, experimental conditions, general predictions, primary measurement constructs, sample size, and exclusion criteria.}

@@

Like Experiment~1, redundant systems are characteristic of the Transmission-only condition, especially in Epoch~III. As the spoken suffixes become more homophonous, the variant spellings are increasingly conditioned on shape rather than color, perhaps because the spoken language signals to learners that the language does not mark color, so they rationalize the system as three words with idiosyncratic spellings. Interestingly, however, all chains exhibited conservation of spelling \textit{form}, even if the way in which spelling was conditioned on meaning was lost. For example, Chain~A ultimately represents the sound /fə/ with the spellings <foe> and <xa>, spellings that are internally inconsistent and contrary to standard uses of the Latin alphabet, but which trace their origins back to the original seed orthography. Overall, then, the Transmission-only condition in Experiment~2 is characterized by the conservation of spelling \textit{form} without conserving how form patterns with meaning.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig12.pdf}}
	\vspace*{2pt}
	\caption{Communicative success by generation in the communicative conditions of all three experiments. The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3). Experiment 3 is a replication of Experiment 1 without any auditory component (see Appendix C in the supplementary material).}
	\label{fig12}
	\end{figure}
